---
title: "Writing and running your first nimble model"
subtitle: "Spatial and spatio-temporal models with NIMBLE"
author: "Tullia Padellini"
date: "1 July 2021"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, cache = TRUE, fig.align = 'center', fig.width = 10)
```

# Writing and running a model in NIMBLE

**N**umerical **I**nference for statistical **M**odels for **B**ayesian and **L**ikelihood **E**stimation (`NIMBLE`)

When the analysis is not conjugate, carrying on Bayesian inference is non-trivial:
since writing your own MCMC can be *complex*, and *time consuming*.
More critically, simulation methods such as MCMC are *computationally very intensive*,
which means that you may need to use programming languages which are faster than `R`,
such as `C++`.

`nimble` is an `R` package that allows to tackle these issues, making it easier 
to fit Bayesian models. The idea behind `nimble` is that the user only needs to 
specify the model, and then `nimble` will automatically write efficient C++ code 
to implement several simulation algorithms. 

As its creators put it:

*One of the most important concepts behind NIMBLE is to allow a combination of high-level processing in R and low-level processing in C++. For example, when we write a Metropolis-Hastings MCMC sampler in the NIMBLE language, the inspection of the model structure related to one node is done in R, and the actual sampler calculations are done in C++.*


##### *Step 1. write the model*

In the first step you need to specify the variables involved in your model and 
how they are related to each other.

Let us consider a basic example where we observe a sample $X_1, \dots, X_n$ of
normally distributed random variables. Mean and variance are common to all 
observations but unknown, hence our goal is to estimate them. 

In a Bayesian fashion, we specify priors for the two unknown parameters of interest, 
thus obtaining the following model:

\begin{align*}
X_i| \mu, \sigma & \sim \text{Normal}(\mu, \sigma) \\
\mu & \sim \text{Normal}(0, 10) \\
\sigma^2 & \sim \text{Uniform}(0,1000)
\end{align*}

Notice that in this case, as we are not working with conjugate priors, it may be 
difficult (if possible at all) to derive the posterior distribution of the parameters
of interest (and the corresponding means), which is why we turn to simulation methods.


In the case of `nimble`, in particular, we need to input only the model as seen 
above, and then the software will automatically produce random draws for the 
posterior distribution.

More specifically, our basic model can be specified as:
```{r eval = FALSE}
code <- nimbleCode({

    # priors: 
    mu ~ dnorm(0, sd = 10)
    sigma2 ~ dunif(0, 1000)
    tau <- 1 / sigma2
    
    # likelihood: 
    for(i in 1:N){
        x[i] ~ dnorm(mu, var = sigma2)
    }
})
```

While the model code may seem similar to `R` code, it is a completely different 
language (more on that later).

</br>


##### *Step 2. compile & run the model*

In the second step you need to specify the basic configuration of your MCMC algorithm 
(i.e. the number of chains you want to run, the number of iterations to be run for each chain and so on)
and what are the data. 

<!-- `nimble` will convert your model specification in `C++`, and then convert the resulting object back to `R` -->

In the case of our toy example: 

```{r, eval = FALSE}
data <- list(x = c(2, 5, 3, 4, 1, 0, 1, 3, 5, 3))

mcmc.out <- nimbleMCMC(code = code, 
                       data = data,
                       nchains = 2, niter = 1000)
```

</br>

## Step 1: Write the model

Writing the model basically consists of specifying:

* **which** are the variables of the problem 
* **how** they are related to each other. 

</br>


In our basic example the *variables* are:

* $X_1, \dots, X_N$: the observations. Each of the $X_i$ is related to $\mu$ and $\sigma$ 
\[X_i| \mu, \sigma  \sim \text{Normal}(\mu, \sigma)\]

* $\mu$: the mean of the ... this is also a random variable 
\[\mu  \sim \text{Normal}(0, 10)\]

* $\sigma^2$
\[\sigma^2 \sim \text{Uniform}(0,1000)\]

* $\tau = 1/\sigma^2$ as a transformation of a random variable, this is still a random variable, 
however, you don't need to specify a distribution for it, it automatically comes from the distribution of $\sigma^2$

</br> 

In general, variables in `nimble` can be of two types: *stochastic* or *deterministic*. 


### **stochastic** variables

Variables on which you make distributional assumptions are called *stochastic*. 
These variables are declared through the symbol `~`, which links the 
name of the variable (e.g. `mu`) to the distribution of your choice (e.g. `dnorm`).

**Example:** 

```{r eval = FALSE}
mu ~ dnorm(0, sd = 1000)
sigma ~ dunif(0, 1000)

for(i in 1:N){
    x[i] ~ dnorm(mu, sd = sigma)
}
```

All most common distribution are already implemented in `nimble`, let us list a 
couple of examples:

* Normal `x ~ dnorm(mu, tau)`
* Poisson `x ~ dpois(lambda)`
* Binomial `x ~ dbinom(n, p)`
* Uniform `x ~ dunif(low, up) `
* Multinomial `x ~ dmulti(p, n)`
* Multivariate normal `x ~ dmnorm(mu, var)`

`nimble` allows to have different parametrization of the same distribution. 
For example in the case of the `normal` distribution, one can choose between:

* `dnorm(mu = 0, sd = 1)` 
* `dnorm(mu = 0, var = 1)`
* `dnorm(mu = 0, tau = 1)`  *default*

If you don't specify the name of the variance parameter, i.e. `dnorm(0,1)`, 
`nimble` will by default consider the second parameter to be the *precision*. 
It's good practice to always explicitly specify the name of the parameter you 
are trying to fix/model. 

Check [here](https://r-nimble.org/html_manual/cha-writing-models.html#tab:distributions)
for a complete list of the distribution in `nimble` and their parametrization. 
You can also define your own probability distribution. 


Multidmensional variables, such as the vector of observations $X = X_1,\dots, X_n$ 
can be represented in `nimble` as multidimensional objects such as vectors or matrices.
The syntax to select one or more elements from a vector or a matrix 
in `nimble` code is the same as basic `R`. 

For example if `x` is a vector, `x[i]` will return its $i$-th entry;
if `x` is a matrix `x[i,j]` will return the element on row $i$ and column $j$. 

In the case of a multidimensional variable, such as the vector of observations 
of our toy example, `x`, we have to specify the distribution of 
all its components. 

This can be done in several ways:

* using a for loop:

```{r eval = FALSE}
for(i in 1:N){
    x[i] ~ dnorm(mu, sd = sigma)
}
```

* joint specification with explicit dimension references:

```{r eval = FALSE}
code <- nimbleCode({
  y[1:K] ~ dmulti(p[1:K], n)
  p[1:K] ~ ddirch(alpha[1:K])
  log(alpha[1:K]) ~ dmnorm(alpha0[1:K], R[1:K, 1:K])
})
```

* joint specification without explicit dimension references:

```{r eval = FALSE}
codeAlt <- nimbleCode({
  y[] ~ dmulti(p[], n)
  p[] ~ ddirch(alpha[])
  log(alpha[]) ~ dmnorm(alpha0[], R[ , ])
})
```

Notice that the following syntax is **invalid**.
```{r, eval = FALSE}
y ~ dmulti(p, n)
```
As opposed to `R`, where you can assign a vector or a scalar with the same syntax,
in `nimble` you always have to refer to the vector structure, whether implicitely
(i.e. using the empty `[]`) or explicitly (i.e. specifying the dimension `[1:K]`)


### **deterministic** variables 

are declared through the symbol `<-`
e.g. we are not interested in the variance $\sigma^2$ but in the precision $\tau = 1/\sigma^2$. 
This is very similar to what you would do in basic `R`.
```{r eval = FALSE}
tau <- 1/ sigma2
```

  
As in the case of stochastic nodes, square brackets must always be provided to indicate number of dimensions in the case of vector variables:
```{r eval = FALSE}
for(i in 1:10) {
    logY[i] <- log(Y[i])
}
```

```{r eval = FALSE}
logY[1:10] <- log(Y[1:10])
```


    
You can also use matrix algebra operations, such as inner product `%*%`: 

```{r eval = FALSE}

```

`nimble` treats vector/matrices differently than scalar.
If `x` is 2-dimensional, use `x[,] %*% beta[]`, not `x %*% beta`

In addition to use `[]` to make clear that you are working with an higher dimensional
object, you need to be careful about scalar vs. vector vs. matrix vs. array. 

This will not work:
```{r eval = FALSE}
x[1:5] <- A[1:5, 1:5] %*% b[1:5] + c[1:5]
```

The problem is that the right-hand-side returns a matrix, so we can't assign it to a vector.
This will work:
```{r eval = FALSE}
x[1:5] <- (A[1:5, 1:5] %*% b[1:5] + c[1:5])[,1]
```


**NB:** not all the functions in `R` are available in `nimble`. Check [here](https://r-nimble.org/html_manual/cha-writing-models.html#tab:functions) for a complete list of functions. 

### Additional code

#### if statements

If you wish to define multiple alternative models in one set of code,
you can use if-then-else statements. These will be evaluated based on
variables in the R environment when the model is defined.  For
example:

```{r eval = FALSE}
code <- nimbleCode({
    sigma ~ dunif(0, 10)
    beta0 ~ dnorm(0, sd = 1000)
    beta1 ~ dnorm(0, sd = 1000)
    if(INCLUDE_X2) { beta2 ~ dnorm(0, sd = 1000) } else {}
    for(i in 1:10) {
        if(INCLUDE_X2) {
            y[i] ~ dnorm(beta0 + beta1 * x1[i] + beta2 * x2[i], sd = sigma)
        } else {
            y[i] ~ dnorm(beta0 + beta1 * x1[i], sd = sigma)
        }
    }
})

INCLUDE_X2 <- FALSE
m1 <- nimbleModel(code)
INCLUDE_X2 <- TRUE
m2 <- nimbleModel(code)
```
Depending on the value of the variable `INCLUDE_X2` different models are chosen.

#### ordering

We have seen before that our basic model:
\begin{align*}
X_i| \mu, \sigma & \sim \text{Normal}(\mu, \sigma) \\
\mu & \sim \text{Normal}(0, 10) \\
\sigma & \sim \text{Uniform}(0,1000)
\end{align*}

Could be translated in `nimble` as follows:
```{r eval = FALSE}
code <- nimbleCode({

    # priors: 
    mu ~ dnorm(0, sd = 1000)
    tau ~ dunif(0, 1000)
    
    sigma2 <- 1 / tau
    
    # likelihood: 
    for(i in 1:10){
        x[i] ~ dnorm(mu, var = sigma2)
    }

})
```



`nimble` is a declarative language, which means that it defines objects but 
is not run "line by line" as it would be the case for basic R. 
Each line declares that a component should be run
but it doesn’t matter in what order they are declared.

The following code is thus completely analogous to the previous one:
```{r eval = FALSE}
code <- nimbleCode({

    # likelihood:
    for(i in 1:10){
        x[i] ~ dnorm(mu, var = sigma2)
    }
    
    # priors
    mu ~ dnorm(0, sd = 1000)
    tau ~ dunif(0, 1000)
    
    sigma2 <- 1 / tau
})
```


## Step 2: running the model

Once you have written the model, the most direct approach to executing an MCMC 
algorithm in NIMBLE is using the function `nimbleMCMC`. This single command can be used to 
create an underlying model and associated MCMC algorithm, compile both of these, 
execute the MCMC, and return samples, summary statistics, and a WAIC value. 


```{r, eval = FALSE}
mcmc.out <- nimbleMCMC(code = code, 
                       constants = list(N=10),
                       data = data, 
                       inits = inits,
                       nchains = 2, 
                       niter = 1000,
                       summary = TRUE,
                       samples = TRUE,
                       monitors = c('mu','sigma2', 'tau'),
                       samplesAsCodaMCMC = TRUE)

```
Let us analyze some of the arguments of `nimbleMCMC`: 

* `code`: a `nimble-code` object, created with the function `nimbleCode`. It contains 
the model specification but no information on the data, hence it can be recycled for 
multiple samples from the same phenomenon. 

* `data`: a list containing the observations to be used in the fitting. The list 
element must have the same name as the corresponding variable in the model specified in `code`, 
so that `nimble` can link them. 
```{r, eval = FALSE}
nimbleData <- list(x = c(2, 5, 3, 4, 1, 0, 1, 3, 5, 3))
```
The separation between the data and the model allows an algorithm to be applied to many datasets in the same model without re-creating the model each time. It also allows simulation of data in a model.

* `constants`: a list containing *structural* variables that do not change in different datasets of the same model, such as for example a vector of known index values, such as for block indices, or variables used in the index ranges of for-loops. Once again, each element of this list must have the same name as the variable in the `code` object it referres to. 

```{r, eval = FALSE}
nimbleConsts = list(N = 10)
```


* `inits` initial values of the Markov Chains. They can be provided either as a list:
```{r, eval = FALSE}
nimbleInits <- list(mu = 1, sigma = 1)
```

either as a function, returning a list:
```{r eval = FALSE}
nimbleInits <- function() list(mu = rnorm(1,0,1), sigma = runif(1,0,10))
```

* `nchains`: number of chains to be run (typically 2 or 4, you can have only 1 as well but multiple chains are required for some of the diagnostics). 
* `thin`: thinning factor
* `niter`: number of MCMC iterations for each chain
* `nburnin`: number of iteration to be discarded as part of the burning in
* `summary`: logical value. If `TRUE`, `nimbleMCMC` will return a summary of the chains it ran
* `samples`: logical value. If `TRUE`, `nimbleMCMC` will return the values generated for each chain, after the burn-in 
* `samplesAsCodaMCMC`: logical value. If `TRUE`, `nimbleMCMC` will return an object of class `coda`. This allows to use the output of the function `nimbleMCMC` in other `R` packages, which is especially relevant for assessing convergence (as we will see later)
* `WAIC`: logical value. If `TRUE`, `nimbleMCMC` will compute and return the WAIC value of the model
* `monitors`: a vector containing the names of the variables/parameters we want `nimble` to produce output for. 
For example if we set
```{r eval =FALSE}
monitors = c('mu','sigma2', 'tau')
```
we will have posterior samples/summaries for $\mu, \sigma^2$ and $\tau$. 
If we set 
```{r eval =FALSE}
monitors = c('mu', 'tau')
```
we will only have samples/summaries for $\mu$ and $\tau$. 


</br>
Let us run our toy model:
```{r, eval = TRUE}
library(nimble)
code <- nimbleCode({

    # likelihood:
    for(i in 1:10)
        x[i] ~ dnorm(mu, var = sigma2)
    
    
    # priors
    mu ~ dnorm(0, sd = 100)
    sigma2 ~ dunif(0, 1000)
    
    tau <- 1 / sigma2
})

data <- list(x = c(2, 5, 3, 4, 1, 0, 1, 3, 5, 3))

initsFunction <- function() list(mu = rnorm(1,0,1), sigma2 = runif(1,0,10))

mcmc.out <- nimbleMCMC(code = code, 
                       constants = list(N=10),
                       data = data, 
                       inits = initsFunction,
                       nchains = 2, 
                       niter = 1000,
                       summary = TRUE,
                       samples = TRUE,
                       monitors = c('mu','sigma2', 'tau'),
                       samplesAsCodaMCMC = TRUE)

```

The object `mcmc.out` is a list containing both the chains we ran and a summary of them:
```{r eval = TRUE}
class(mcmc.out)
names(mcmc.out)
```
Let's have a quick look. The object `$samples` is another list, containing the random 
draws for each iteration of the MCMC sampler:
```{r eval = TRUE}
head(mcmc.out$samples)
```


The `$summary` is also a list, whose entries are a summary of the posterior 
distribution of the parameters monitored:
```{r}
mcmc.out$summary
```

As you can see the point estimate $\hat{\tau}$ is different than $1/\hat{\sigma}$, 
which is why if you are interested in some transformation of the parameter is convenient 
to define such tranformation in your model specification and then monitor it.

